---
title: "Complex Data - lab4"
author: "Stanisław Wilczyński"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ALA)
library(latex2exp)
library(ggplot2)
library(reshape2)
library(nlme)
library(lattice)
library(magrittr)
library(knitr)
library(kableExtra)
```


```{r}
hrunbalanced <- read.csv(file = "../data/hrunbalanced.csv", header = TRUE)
means <- tapply(hrunbalanced$hr, list(hrunbalanced$time, hrunbalanced$drug), mean)
xyplot(hr~time|factor(drug), type='l',groups=id,data=hrunbalanced, main = "Spaghetti plot of the heart rate over time in 2 groups")
```
```{r}
## Random intercept
fit.RandInt <- lme(hr ~ time*factor(drug),
random=~1|id,
data=hrunbalanced)

#Random slope
fit.RandSlp <- lme(hr ~ time*factor(drug),
random=~1+time|id,
data=hrunbalanced)

covar <- getVarCov(fit.RandSlp)
varCorr <- VarCorr(fit.RandSlp)
```

## Task 1

Here we just perform proper computations, as in the task description. First we calculate the means.
```{r}
#Calculating matrices (X,Z)
coefs <- fit.RandSlp$coefficients$fixed
coefs <- matrix(c(coefs[1], coefs[3], coefs[2], coefs[4]), ncol = 1)
ones <- matrix(rep(1,6), nrow=6)
zeros <- matrix(rep(0,6), nrow = 6)
tN <- matrix(c(0,1,5,15,30,60), nrow = 6)
t1 <- matrix(hrunbalanced$time[1:6], nrow = 6)
t2 <- matrix(hrunbalanced$time[7:12], nrow = 6)
t13 <- matrix(hrunbalanced$time[73:76], nrow = 4)
x1 <- cbind(ones,zeros,t1,zeros)
x2 <- cbind(ones,zeros, t2,zeros)
x13 <- cbind(ones[1:4,],ones[1:4,],t13,t13)
xNA <- cbind(ones, zeros, tN, zeros)
xNB <- cbind(ones, ones, tN, tN)
ZN <- cbind(ones,tN)
Z1 <- cbind(ones,t1)
Z2 <- cbind(ones, t2)
Z13 <- cbind(ones[1:4], t13)
```

```{r, results='asis'}
#Calculating means
mA <- round(xNA%*%coefs,1)
mB <- round(xNB%*%coefs,1)
m1 <- round(x1%*%coefs,1)
m2 <- round(x2%*%coefs,1)
m13 <- rbind(round(x13%*%coefs,1), c(NA), c(NA))
colnames(mA) <- c("Mean (drug A)")
colnames(mB) <- c("Mean (drug B)")
colnames(m1) <- c("Mean (obs. 1)")
colnames(m2) <- c("Mean (obs. 2)")
colnames(m13) <- c("Mean (obs. 13)")

kable(cbind(mA,m1,m2,mB,m13), format = "latex", booktabs=TRUE, 
          caption = "Means") %>% kable_styling(latex_options="HOLD_position")
```

And now we calculate the covariances:
```{r}
getCov <- function(Z){
  Z%*%covar%*%t(Z) + as.numeric(varCorr[3,1])*diag(1,nrow = dim(Z)[1], ncol = dim(Z)[1])
}
#Calucalting covariances
covZN <- round(getCov(ZN),1)
covZ1 <- round(getCov(Z1),1)
covZ2 <- round(getCov(Z2),1)
covZ13 <- round(getCov(Z13),1)
colnames(ZN) <- c("Cov. (standard)","")
colnames(Z1) <- c("Cov. (obs. 1)","")
colnames(Z2) <- c("Cov. (obs. 2)","")
colnames(Z13) <- c("Cov. (obs. 13)","")


kable(covZN, format = "latex", booktabs=TRUE, 
      caption = "Cov. (standard)") %>% kable_styling(latex_options="HOLD_position")
kable(covZ1, format = "latex", booktabs=TRUE, 
      caption = "Cov. (obs. 1)") %>% kable_styling(latex_options="HOLD_position")
kable(covZ2, format = "latex", booktabs=TRUE, 
      caption = "Cov. (obs. 2)") %>% kable_styling(latex_options="HOLD_position")
kable(covZ13, format = "latex", booktabs=TRUE, 
      caption = "Cov. (obs. 13)") %>% kable_styling(latex_options="HOLD_position")
```

## Task 2

```{r}
## step one
coef.hr <- matrix(NA,ncol = 3, nrow = 0)
for (i in unique(hrunbalanced$id)) {
  model <- lm(hr~time, data = hrunbalanced, subset = id == i)
  model.drug <- unique(hrunbalanced[hrunbalanced$id == i, "drug"])
  coef.hr <- rbind(coef.hr, c(model.drug, coef(model)))
}
colnames(coef.hr) <- c("drug","intercept","slope")
coef.hr <- as.data.frame(coef.hr)
print(coef.hr)
```

```{r}

## step two
hr.int <- lm(intercept~drug, data = coef.hr)
hr.slp <- lm(slope~drug, data=coef.hr)

sum.int <- summary(hr.int)
sum.slp <- summary(hr.slp)
sum.lme <- summary(fit.RandSlp)
print(sum.int)
print(sum.slp)

info.int <- round(sum.int$coefficients[,c(1,4)], 4)
info.slp <- round(sum.slp$coefficients[,c(1,4)], 4)
info.mixed <- round(summary(fit.RandSlp)$tTable[,c(1,5)], 4)
```
From above summaries we can clearly see that p-values for $drug$ effect are greater than $0.05$. Therefore, we conclude that neither the mean intercepts, nor the mean slopes differ significantly in the two groups.

Below we provide a table with coefficients' estimates and their p-values bot in Two-Stage and Mixed Effects analysis.

\begin{center}
  \begin{tabular}{c c c}
     & Two-Stage & Mixed Effects \\ \hline
    Intercept & $`r info.int[1,1]` (`r info.int[1,2]`)$ & $`r info.mixed[1,1]` (`r info.mixed[1,2]`)$ \\
    Time & $`r info.slp[1,1]` (`r info.slp[1,2]`)$ & $`r info.mixed[2,1]` (`r info.mixed[2,2]`)$\\
    Drug & $`r info.int[2,1]` (`r info.int[2,2]`)$ & $`r info.mixed[3,1]` (`r info.mixed[3,2]`)$ \\
    Time*Drug & $`r info.slp[2,1]` (`r info.slp[2,2]`)$ & $`r info.mixed[4,1]` (`r info.mixed[4,2]`)$ \\
    \hline
  \end{tabular}
\end{center}

We can see that there are small differences both in estimates and the p-values when comparing these two models, however the interpretation didn't change - in both models only $time$ and $intercept$ are significant and they both indicate that the drug effect is not significant.
