---
title: "Complex data - project"
author: "Stanisław Wilczyński, Anna Zaleska"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#knitr::opts_chunk$set(results = 'asis');
options(xtable.comment=FALSE)
library(PBImisc)
library(xtable)
library(lattice)
library(lme4)
library(nlme)
library(neuralnet)
#library(lmerTest)

data(kidney)
```


```{r}

xy.kidney <- reshape(kidney, direction = "long", varying = list(9:16), v.names = c("MDRD"), times = c(1/4,1,3,6,12,24,36,60))
xy.kidney$ltime <- log(xy.kidney$time)

square.panel <- function(...) {
  nx = list(...)$x
  ny = list(...)$y
  cc = lm(ny~nx+I(nx^2))$coef
  panel.xyplot(...)
  panel.curve(x^2*cc[3]+x*cc[2]+cc[1], from = min(nx), to=max(nx), col = "red")
}

xyplot(MDRD ~ time | id, xy.kidney, type=c("g","p","r"), xlab = "time since transplant", ylab = "MDRD", pch=19, layout=c(3,4), panel = square.panel, subset = id>150 & id<163)
xyplot(MDRD ~ ltime | id, xy.kidney, type=c("g","p","r"), xlab = "log(time since transplant)", ylab = "MDRD", pch=19, layout=c(3,4), panel = square.panel, subset = id>150 & id<163)

lkidney <- reshape(kidney, direction = "long", varying = list(9:16), v.names = c("MDRD"), times = c(1/4,1,3,6,12,24,36,60))
lkidney$ltime <- log(lkidney$time)

colMeans(kidney[,9:16])
```


### Fixed effects

```{r}
for(i in 1:8){
  model.variable <- lm(MDRD ~ lkidney[,i], data = lkidney)
  print(paste("Model for", colnames(lkidney)[i]))
  print(summary(model.variable))
  print("---------------------------------------------------------------------------")
  #donor.age, CIT, discrepanacy.DR
}
```



## time as factor
```{r}
un.model.reml <- gls(MDRD~donor.age+bpl.drugs+recipient.age+discrepancy.AB+discrepancy.DR+CIT+factor(time)*(diabetes+therapy),
correlation=corSymm(form= ~1 | id),
weights=varIdent(form= ~1 | time),
data=lkidney)

cs.model.reml <- gls(MDRD~donor.age+bpl.drugs+recipient.age+discrepancy.AB+discrepancy.DR+CIT+factor(time)*(diabetes+therapy),
correlation=corCompSymm(form= ~1 | id),
weights=varIdent(form= ~1),
data=lkidney)

ar.model.reml <- gls(MDRD~donor.age+bpl.drugs+recipient.age+discrepancy.AB+discrepancy.DR+CIT+factor(time)*(diabetes+therapy),
correlation=corAR1(form= ~1 | id),
weights=varIdent(form= ~1),
data=lkidney)

car.model.reml <- gls(MDRD~donor.age+bpl.drugs+recipient.age+discrepancy.AB+discrepancy.DR+CIT+factor(time)*(diabetes+therapy),
correlation=corCAR1(form= ~ltime | id),
data=lkidney)

```



```{r}
anova(un.model.reml,ar.model.reml, cs.model.reml, car.model.reml)
anova(un.model.reml)
```
```{r}
un.model <- gls(MDRD~donor.age+bpl.drugs+recipient.age+discrepancy.AB+discrepancy.DR+CIT+factor(time)*(diabetes+therapy),
correlation=corSymm(form= ~1 | id),
weights=varIdent(form= ~1 | time),
data=lkidney, method = "ML")

un.model.no.int <- gls(MDRD~factor(time)+(donor.age+bpl.drugs+recipient.age+discrepancy.AB+diabetes+discrepancy.DR+CIT+therapy),
correlation=corSymm(form= ~1 | id),
weights=varIdent(form= ~1 | time),
data=lkidney, method = "ML")

quad.model <- gls(MDRD~(ltime + I(ltime^2))*(therapy+diabetes) + (donor.age+bpl.drugs+recipient.age+discrepancy.AB+discrepancy.DR+CIT),
correlation=corSymm(form= ~1 | id),
weights=varIdent(form= ~1 | time),
data=lkidney, method = "ML")

lin.model <- gls(MDRD~(ltime)*(therapy+diabetes) + (donor.age+bpl.drugs+recipient.age+discrepancy.AB+discrepancy.DR+CIT),
correlation=corSymm(form= ~1 | id),
weights=varIdent(form= ~1 | time),
data=lkidney, method = "ML")

cube.model <- gls(MDRD~(ltime + I(ltime^2)+I(ltime^3))*(therapy+diabetes) + (donor.age+bpl.drugs+recipient.age+discrepancy.AB+discrepancy.DR+CIT),
correlation=corSymm(form= ~1 | id),
weights=varIdent(form= ~1 | time),
data=lkidney, method = "ML")
```


```{r}
anova(quad.model, un.model)
anova(quad.model, lin.model)
anova(un.model.no.int, un.model)
anova(un.model, cube.model)
anova(un.model)
```

```{r}

un.model <- gls(MDRD~diabetes+donor.age+bpl.drugs+recipient.age+discrepancy.AB+discrepancy.DR+CIT+factor(time)*(therapy),
correlation=corSymm(form= ~1 | id),
weights=varIdent(form= ~1 | time),
data=lkidney, method = "ML")

simpler.un.model <- gls(MDRD~factor(time)*(therapy+diabetes) + (donor.age+bpl.drugs),
correlation=corSymm(form= ~1 | id),
weights=varIdent(form= ~1 | time),
data=lkidney, method = "ML")

simpler.un.model2 <- gls(MDRD~factor(time)*(therapy) + (donor.age+bpl.drugs+diabetes),
correlation=corSymm(form= ~1 | id),
weights=varIdent(form= ~1 | time),
data=lkidney, method = "ML")

simplest.un.model <- gls(MDRD~factor(time)*(therapy) + (donor.age+bpl.drugs),
correlation=corSymm(form= ~1 | id),
weights=varIdent(form= ~1 | time),
data=lkidney, method = "ML")

```

```{r}
anova(simpler.un.model, un.model)
anova(simpler.un.model, simplest.un.model)
anova(simplest.un.model, un.model)
anova(simpler.un.model2, simplest.un.model)
anova(simpler.un.model2, un.model)
anova(simpler.un.model2)
anova(un.model)
anova(simpler.un.model)
```
```{r}
xtable(matrix(c(summary(simplest.un.model)$coefficients, 1,1,1,1), nrow=5, byrow = T), digits = 2, caption = "Coefficients")
```



```{r}
simplest.un.model.lin <- gls(MDRD~ltime+therapy+donor.age+bpl.drugs,
correlation=corSymm(form= ~1 | id),
weights=varIdent(form= ~1 | time),
data=lkidney, method = "ML")

simplest.un.model.quad <- gls(MDRD~(ltime+I(ltime^2))+therapy+donor.age+bpl.drugs,
correlation=corSymm(form= ~1 | id),
weights=varIdent(form= ~1 | time),
data=lkidney, method = "ML")

simplest.un.model.cube <- gls(MDRD~(ltime+I(ltime^2)+I(ltime^3))+therapy+donor.age+bpl.drugs,
correlation=corSymm(form= ~1 | id),
weights=varIdent(form= ~1 | time),
data=lkidney, method = "ML")
```

```{r}
anova(simplest.un.model, simplest.un.model.quad, simplest.un.model.cube)
```

### Prediction

```{r}
explained.var <- function(model, data.set) {
  var1 <- var(data.set$MDRD)
  var2 <- var(residuals(model))
  (1-var2/var1)*100
}

analyse.model <- function(model, train.data, test.data){
  model.test.pred <- predict(model, test.data)
  model.train.pred <- predict(model, train.data)
  model.train.mse <- mean((model.train.pred-train.data$MDRD)**2) #the same as residuals
  model.test.mse <- mean((model.test.pred-train.data$MDRD)**2)
  c(model.train.mse, model.test.mse)
}

analyse.models <- function(data.set=kidney, normalize=F, data.split = 0.8){
  smp.size <- floor(data.split * nrow(data.set))
  train.ind <- sample(seq_len(nrow(data.set)), size = smp.size)
  nkidney <- data.set
  ntime <- log(c(1/4,1,3,6,12,24,36,60))
  if(normalize) {
    for(i in (1:ncol(data.set))){
      if(is.numeric(data.set[2,i])){
        nkidney[,i] <- scale(data.set[,i])
      } 
    }
    ntime <- scale(ntime)
  }
  # changing to long representation 
  kidney.train <- nkidney[train.ind, ]
  kidney.test <- nkidney[-train.ind, ]
  
  lkidney.train <- reshape(kidney.train, direction = "long", varying = list(9:16), v.names = c("MDRD"), times = ntime)
  
  lkidney.test <- reshape(kidney.test, direction = "long", varying = list(9:16), v.names = c("MDRD"), times = ntime)
  
  # fitting models
  
  n.growth.model <- lme(MDRD~(time + I(time^2))*therapy  + donor.age +  bpl.drugs , random =~time + I(time^2)|factor(id), data = lkidney.train)
  
  n.growth.model.cube <- lme(MDRD~(time + I(time^2)+I(time^3))*therapy  + donor.age +  bpl.drugs , random =~time + I(time^2)+I(time^3)|factor(id), data = lkidney.train)
  

  n.simplest.un.model <- gls(MDRD~factor(time)*therapy+donor.age+bpl.drugs,
  correlation=corSymm(form= ~1 | id),
  weights=varIdent(form= ~1 | time),
  data=lkidney.train)
  
  n.un.full.model <- gls(MDRD~donor.age+bpl.drugs+recipient.age+discrepancy.AB+discrepancy.DR+CIT+factor(time)*(diabetes+therapy),
  correlation=corSymm(form= ~1 | id),
  weights=varIdent(form= ~1 | time),
  data=lkidney.train)
  
  n.simplest.quad.model <- gls(MDRD~(time+I(time^2))*therapy+donor.age+bpl.drugs,
  correlation=corSymm(form= ~1 | id),
  weights=varIdent(form= ~1 | time),
  data=lkidney.train)
  
  all.models <- list(n.growth.model, n.growth.model.cube, n.simplest.un.model, n.simplest.quad.model, n.un.full.model)
  
  AICs <- anova(n.growth.model, n.growth.model.cube, n.simplest.un.model, n.simplest.quad.model, n.un.full.model)$AIC
  vars <- sapply(all.models, explained.var, rbind(lkidney.train, lkidney.test))
  errors <- sapply(all.models, analyse.model, lkidney.train, lkidney.test)
  result <- rbind(AICs, vars, errors)
  result <- as.data.frame(result)
  rownames(result) <- c("AIC", "% var.", "Train err.", "Test err.")
  if(!normalize){
    nkidney <- data.set
    ntime <- log(c(1/4,1,3,6,12,24,36,60))
    for(i in (1:ncol(data.set))){
      if(is.numeric(data.set[2,i])){
        nkidney[,i] <- scale(data.set[,i])
      } 
    }
    ntime <- scale(ntime)
    # changing to long representation 
    kidney.train <- nkidney[train.ind, ]
    kidney.test <- nkidney[-train.ind, ]
    
    lkidney.train <- reshape(kidney.train, direction = "long", varying = list(9:16), v.names = c("MDRD"), times = ntime)
    
    lkidney.test <- reshape(kidney.test, direction = "long", varying = list(9:16), v.names = c("MDRD"), times = ntime)
  }
  f <- as.formula("MDRD~donor.age+bpl.drugs+time")
  nn <- neuralnet(f, data=lkidney.train, hidden=c(9,3), threshold = 0.1, linear.output=T)
  pr.nn.test <- as.vector(compute(nn, lkidney.test[,c(2, 8, 9)])$net.result)
  real.values.test <- lkidney.test$MDRD
  if(!normalize){
    real.values.test <- lkidney.test$MDRD*sd(lkidney$MDRD-mean(lkidney$MDRD))+mean(lkidney$MDRD)
    pr.nn.test <- pr.nn.test*sd(lkidney$MDRD-mean(lkidney$MDRD))+mean(lkidney$MDRD)
  }
  test.err <- mean((real.values.test-pr.nn.test)**2)
  pr.nn.train <- as.vector(compute(nn, lkidney.train[,c(2, 8, 9)])$net.result)
  real.values.train <- lkidney.train$MDRD
  if(!normalize){
    pr.nn.train <- pr.nn.train*sd(lkidney$MDRD-mean(lkidney$MDRD))+mean(lkidney$MDRD)
    real.values.train <- lkidney.train$MDRD*sd(lkidney$MDRD-mean(lkidney$MDRD))+mean(lkidney$MDRD)
  }
  train.err <- mean((real.values.train-pr.nn.train)**2)
  result <- cbind(result, c(NA, NA, train.err, test.err))
  colnames(result) <- c("Mixed(q.)", "Mixed(c.)", "Fixed(s.)", "Fixed(q.)", "Fixed(f.)", "NNet")
  result
}
```
```{r}
mean((predict(n.growth.model, lkidney.test, level = 1)-lkidney.test$MDRD)**2)
mean((predict(n.growth.model, lkidney.train, level = 1)-lkidney.train$MDRD)**2)
mean((predict(n.un.full.model, lkidney.train)-lkidney.train$MDRD)**2)
mean((predict(n.un.full.model, lkidney.test)-lkidney.test$MDRD)**2)
analyse.model(n.un.full.model, lkidney.train, lkidney.test)
```



```{r}
#for debugging analyse.models
data.split=0.8
normalize=F
data.set=kidney
#plot(nn, rep = "best", information = F, show.weights = F)
```


```{r}
result.list <- list()
for(i in 1:5){
  print(paste(i))
  result.list[[i]] <- analyse.models()
}
res <- (result.list[[1]]+result.list[[2]]+result.list[[3]]+result.list[[4]]+result.list[[5]])/5
xtable(res)
```



```{r}
result.list.norm <- list()
for(i in 1:5){
  result.list.norm[[i]] <- analyse.models(normalize = T)
}
```


### Mixed effects models



#### random intercept
```{r}
const.mixed.model2 <- lme(MDRD~ltime, random =~1|factor(id),
                         data = lkidney, method = "ML")
summary(const.mixed.model2)
```


####random intercept and slope
```{r}
lin.mixed.model2 <- lme(MDRD~ltime, random =~ltime|factor(id),
                         data = lkidney, method = "ML")
summary(lin.mixed.model2)
```


#### random intercept and slope quadratic model

```{r}

quadr.mixed.model2 <- lme(MDRD~ltime + I(ltime^2), random =~ltime + I(ltime^2)|factor(id),
                         data = lkidney, method = "ML")
summary(quadr.mixed.model2) 

```


```{r}
anova(const.mixed.model2,lin.mixed.model2,quadr.mixed.model2)
```


#### full model
```{r}
full.model2 <- lme(MDRD~donor.age + recipient.age + therapy + diabetes + bpl.drugs + discrepancy.AB + discrepancy.DR + ltime + I(ltime^2)  , random =~ltime + I(ltime^2)|factor(id), data = lkidney,  method = "ML")
summary(full.model2)
```


```{r}
final.model <- lme(MDRD~donor.age + bpl.drugs + ltime + I(ltime^2)  , random =~ltime + I(ltime^2)|factor(id), data = lkidney,  method = "ML")
final.model2 <- lme(MDRD~donor.age + therapy +  bpl.drugs + ltime + I(ltime^2)  , random =~ltime + I(ltime^2)|factor(id), data = lkidney,  method = "ML")
final.model3 <- lme(MDRD~donor.age + therapy +  bpl.drugs + recipient.age + ltime + I(ltime^2)  , random =~ltime + I(ltime^2)|factor(id), data = lkidney,  method = "ML")
anova(final.model, final.model2)
anova(final.model2, final.model3)
```


```{r}
library(car)
qqPlot(ranef(final.model2)[,1], envelope = FALSE, ylab = expression(a[i]))
qqPlot(ranef(final.model2)[,1], envelope = FALSE, ylab = expression(b[i]))
qqPlot(ranef(final.model2)[,1], envelope = FALSE, ylab = expression(c[i]))
```

#### two stage model - therapy interaction

```{r}
lkidney$ltime2 <- lkidney$ltime^2
coef.kidney <- matrix(NA,ncol = 7, nrow = 0)
for (i in unique(lkidney$id)) {
  model <- lm(MDRD~ltime+ltime2, data = lkidney, subset = id == i)
  model.age <- unique(lkidney[lkidney$id == i, "donor.age"])
  model.drug <- unique(lkidney[lkidney$id == i, "bpl.drugs"])
  model.ther <- unique(lkidney[lkidney$id == i, "therapy"])
  coef.kidney <- rbind(coef.kidney, c(i, model.age, model.drug, model.ther, coef(model)))
  #print(coef(model))
}
colnames(coef.kidney) <- c("id","donor.age","bpl.drugs","therapy","intercept","slope", "quadr.slope")
coef.kidney <- as.data.frame(coef.kidney)
#coef.kidney
```




```{r}
### step two
kidney.int <- lm(intercept~factor(therapy)+donor.age+bpl.drugs, data = coef.kidney)
kidney.slp <- lm(slope~factor(therapy), data = coef.kidney)
kidney.quadr.slp <- lm(quadr.slope~factor(therapy), data = coef.kidney)

info.int <- round(summary(kidney.int)$coefficients[,c(1,4)], 4)
info.int
info.slp <- round(summary(kidney.slp)$coefficients[,c(1,4)], 4)
info.quadr <-round(summary(kidney.quadr.slp)$coefficients[,c(1,4)], 4)

summary(kidney.int)$coefficients
```

```{r}

growth.model <- lme(MDRD~(ltime + I(ltime^2))*therapy  + donor.age +  bpl.drugs , random =~ltime + I(ltime^2)|factor(id), data = lkidney)
info.mixed <- round(summary(growth.model)$tTable[,c(1,5)], 4)
summary(growth.model)$tTable
```


\begin{center}
  \begin{tabular}{c c c}
     & Two-Stage & Mixed Effects \\ \hline
    Intercept & $`r info.int[1,1]` (`r info.int[1,2]`)$ & $`r info.mixed[1,1]` (`r info.mixed[1,2]`)$ \\
    logTime & $`r info.slp[1,1]` (`r info.slp[1,2]`)$ & $`r info.mixed[2,1]` (`r info.mixed[2,2]`)$\\
    $logTime^2$ & $`r info.quadr[1,1]` (`r info.quadr[1,2]`)$ & $`r info.mixed[3,1]` (`r info.mixed[3,2]`)$ \\
   donor.age & $`r info.int[2,1]` (`r info.int[2,2]`)$ & $`r info.mixed[4,1]` (`r info.mixed[4,2]`)$ \\
    therapy.cm & $`r info.int[3,1]` (`r info.int[3,2]`)$ & $`r info.mixed[5,1]` (`r info.mixed[5,2]`)$ \\
   therapy.tc & $`r info.int[4,1]` (`r info.int[4,2]`)$ & $`r info.mixed[6,1]` (`r info.mixed[6,2]`)$ \\
      bpl.drugs & $`r info.int[5,1]` (`r info.int[5,2]`)$ & $`r info.mixed[7,1]` (`r info.mixed[7,2]`)$ \\
     logTime*therapy.cm & $`r info.slp[2,1]` (`r info.slp[2,2]`)$ & $`r info.mixed[8,1]` (`r info.mixed[8,2]`)$ \\
    logTime*therapy.tc & $`r info.slp[3,1]` (`r info.slp[3,2]`)$ & $`r info.mixed[9,1]` (`r info.mixed[9,2]`)$ \\
     $logTime^2$*therapy.cm & $`r info.quadr[2,1]` (`r info.quadr[2,2]`)$ & $`r info.mixed[10,1]` (`r info.mixed[10,2]`)$ \\
    $logTime^2$*therapy.tc & $`r info.quadr[3,1]` (`r info.quadr[3,2]`)$ & $`r info.mixed[11,1]` (`r info.mixed[11,2]`)$ \\

    \hline
  \end{tabular}
\end{center}


```{r}
growth.model <- lme(MDRD~(ltime + I(ltime^2))*therapy  + donor.age +  bpl.drugs , random =~ltime + I(ltime^2)|factor(id), data = lkidney)

var.MDRD <- var(lkidney$MDRD)
var.growth.model <- var(residuals(growth.model))
(1-var.growth.model/var.MDRD)*100
```

#### ltime vs therapy interaction
```{r}
final.model2.ml.inter <- lme(MDRD~donor.age + therapy +  bpl.drugs + ltime + I(ltime^2) + ltime*therapy  , random =~ltime + I(ltime^2)|factor(id), data = lkidney, method = "ML")
anova(final.model2.ml.inter, final.model2)
```
```{r}
growth.model.ml <- lme(MDRD~(ltime + I(ltime^2))*therapy  + donor.age +  bpl.drugs , random =~ltime + I(ltime^2)|factor(id), data = lkidney, method = "ML")
anova(final.model2.ml.inter,growth.model.ml)
```


### cubic time
```{r}
final.model3.ml <- lme(MDRD~donor.age + therapy +  bpl.drugs + ltime + I(ltime^2) +I(ltime^3)  , random =~ltime + I(ltime^2) + I(ltime^3)|factor(id), data = lkidney, method = "ML")
anova(final.model2, final.model3.ml )
```



